---
title: "Lecture 5. ggplot(3)"
author: "Heesun Yoo"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(ggplot2)
library(ggthemes)
library(tidyverse)
```

##exercise
Using mpg data, Create a scatter plot with 'displ' (engine displacement) on the x-axis and 'hwy' (highway miles per gallon) on the y-axis. Coloring based on 'displ' of the car using gradient color scheme. And split the scatter plot into multiple plots (facets) based on the cyl variable. Add the label of manufacturer on each point. 

```{r}
ggplot(mpg, aes(displ, hwy))+
  geom_point(aes(color=displ))+
  scale_color_gradient(name="Engine Displacement", low="magenta", high="blue")+
  facet_grid(~cyl)+
  geom_text(aes(label=manufacturer), color="black", size=2, check_overlap=T)
```

##excercise
Using midwest data, Create a scatter plot with 'popdensity' on the x-axis and 'percollege' on the y-axis. Coloring based on 'popasian' using gradient color scheme, using rainbow(5). Add the geom_smooth tracing the general trend. Add the label of state on each point. 

```{r}
ggplot(midwest, aes(popdensity, percollege))+
  geom_point(aes(color=popasian))+
  scale_color_gradientn(name="population asian", colors=rainbow(5))+
  geom_smooth()+
  geom_text(aes(label=state),color="black", size=1.5, check_overlap=T)
```

So far, the ggplot object that we have learned are mainly scatter plots, where x and y axis variables are numeric and continuous variable. In this lecture, we are going to learn how to visualize the categorical variables with ggplot. 
 
## barplot in ggplot
: represent frequencies of different categories within a categorical variable 

One variable 
```{r}
# barplot in ggplot : one categorical variable - no variable in y axis, why?
ggplot(diamonds, aes(x=cut))+
  geom_bar(fill="blue", color="black")+
  theme_bw()

 #fill is for the filling color, color is for the borderline color
```

Two variables 
```{r}
# barplot in ggplot : add variables by mapping the variable to the aesthetic attribute. 

# rainbow(n) to generate a vector of n colors
# For each bar that represents each cut category, we will segment this bar using the different fill colors of 'clarity' categories. 

# when choosing the fill color to represent each clarity category, we use scale_fill_manual 
 
ggplot(diamonds, aes(x=cut))+
  geom_bar(aes(fill=clarity), color="black")+
  scale_fill_manual(values=rainbow(8))+ 
  theme_bw()

table(diamonds$clarity)

```

```{r}
# position="dodge" 
ggplot(diamonds, aes(x=cut))+
  geom_bar(aes(fill=clarity), position="dodge")+
  scale_fill_manual(values=rainbow(8))+
  theme_bw()
```




## boxplot in ggplot
Categorical variable on x axis, and quantitative variable on y axis

```{r}
# boxplot in ggplot: 
ggplot(diamonds, aes(x=cut, y=price))+
  geom_boxplot(fill=terrain.colors(5), color="darkblue", alpha=0.8)+ 
  theme_bw()
# we did not use the aesthetic mapping here (did not incorporate 3rd variable) 
# cm.colors: Generates n cyan-to-magenta colors.
```

## exercise
Using the titanic dataset, create a bar plot showing the frequencies of the 'Survived' variable, with each bar divided by 'Class' categories. Assign different colors to these categories, and display the bars side-by-side
```{r}
titanic <- as.data.frame(Titanic)
```


## exercise
Using the iris dataset, visualize the relationship between "Species" (a categorical variable) and "Petal.Length" (a quantitative variable). Select the appropriate type of plot and generate it.

```{r} 
library(tidyverse)
table(iris$Species)
```

## histogram in ggplot

```{r}
#using R basic graph      
hist(diamonds$price)

## one quantitative variable
ggplot(diamonds, aes(x=price))+
  geom_histogram(bins=10, fill=topo.colors(10), color="black", alpha=0.8)


# histogram's y axis is the frequency corresponding to a given range of x values. We don't need to specify y variable.

# we did not use the aesthetic mapping here but you can use it. 
#topo.colors(n):Provides n colors representing topographic elevation
```



## density, line graphs in ggplot

```{r}
#density: smoothed version of the histogram
ggplot(diamonds, aes(x=price))+ # y axis represents the probability density.
  geom_density() 

## add lines to time series data
# ?economics 
ggplot(economics, aes(x=date, y= unemploy))+
  geom_line()+
  labs(x="date", y="number of unemployed in thousands")+
  geom_vline(xintercept = as.Date("2010-01-01"), color="red")

class(economics$date)

```


## Statistical transformations

```{r}
# geom_bar(stat="count") is the default: counting the frequency of each category  
ggplot(diamonds, aes(x=cut))+
  geom_bar()  

ggplot(diamonds, aes(x=cut))+
  stat_count() 
```


What if we want to display the proportion that each category accounts for instead of raw frequencies on y axis?
```{r}
## Basic R Graph
barplot(prop.table(table(diamonds$cut)), ylim=c(0,0.5))

```

```{r}
## barplot in ggplot 

#first convert the proportion table as a dataframe
a <- prop.table(table(diamonds$cut)) # a is a table 
a

b <- as.data.frame(a) # it is possible to turn a table into a dataframe
names(b) <- c("cut", "prop") # upon conversion, change the variable's name
b

ggplot(b, aes(x=cut, y=prop))+
  geom_bar(stat="identity")  #When you have pre-computed values (in our context, "prop") and you want the heights of the bars to represent these values directly, you use stat = "identity".   

#
ggplot(b, aes(x=cut, y=prop))+
  geom_col()  # geom_bar(stat="identity") is the same as geom_col()

#
ggplot(b, aes(x=cut, y=prop))+
  geom_col()+
  coord_flip() # can rotate the barplot
```


